\Ace: Growing a Statically-Typed Language Inside a Python

### 2013-11-14 ace-pldi14/ace-pldi14.tex
pldi14 2.5 p3+4+5

The type \verb|TF| on line \ref{compscript}.5 is extracted from the generic function \verb|negate| constructed in Listing \ref{metaprogramming}. Generic functions, according to Sec. \ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless assigning a type to it.

This notion of types as metalanguage objects is key to the Ace compilation model and also enables other mechanisms that we will discuss in subsequent sections.

\subsection{Type Propagation}
The type assigned to the third argument, \verb|f|, on both Lines 4.9 and 4.10, is \verb|add5.ace_type|. The \verb|ace_type| attribute of a generic function is an instance of \verb|ace.GenericFnType|, the type of Ace generic functions. Ace generic functions are compiled to concrete functions automatically at all internal call sites. That is, when the compiler encounters the call to \verb|f| inside \verb|map| when compiling \verb|map_add5_double|, it compiles a version of \verb|add5| specialized to the \verb|double| type (seen on Line 5.3), and similarly when compiling \verb|map_add5_int| (on Line 5.16, automatically given a unique name to avoid conflicts). This mechanism is called {\em type propagation}. We did not need to use \verb|add5.compile(double)| before compiling \verb|map_add5_dbl| because only functions that are never called in the process of compiling other functions in a module need type information explicitly provided, supporting \textbf{ease-of-use} by increasing conciseness.

In effect, this scheme allows for a form of higher-order functional programming even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers). This works because the \verb|ace.GenericFnType| for one function, such as \verb|add5|, is not equal to  the \verb|ace.GenericFnType| for a superficially similar function, such as \verb|add6| (defined as one would expect). To put it in type theoretic terms, \verb|ace.GenericFnType|s are singleton types, {\em uniquely inhabited} by a single generic function. A consequence of this is that they are not useful as first-class values (i.e. they cannot be written into a collection). This is often valuable, particularly in parallel programming where compile-time specialization is valuable to avoid \textbf{performance} and \textbf{ease-of-use} issues that occur when using function pointers.

Concrete functions, on the other hand, can be given a true function type (e.g. \verb|add5| could be compiled to a concrete function with type \verb|int| $\rightarrow$ \verb|int|) if targeting a backend that supports them, such as C99, or by using an integer-indexed jump table in OpenCL (we have not implemented this mechanism using Ace as of the time of writing, but do not anticipate difficulties in doing so).

Type propagation via generic functions can be compared to template specialization in C++, where both the template headers (containing nested template parameters for function arguments) and specialization parameters at any call sites are inferred automatically from usage. This significantly simplifies a sophisticated feature of C++ and introduces it to OpenCL and C, which do not support templates. Other uses for C++ templates are subsumed by the metaprogramming features discussed in Section 5.
% Is this really any more concise than using C + templates? Etc. relation to types. Conciseness.

\subsection{Whole-Function Type Inference}
\begin{codelisting}
\lstinputlisting{listing8.py}
\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
\label{py}
\end{codelisting}

%\begin{codelisting}
%\lstinputlisting{listing6.py}
%\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
%\label{inference}
%\end{codelisting}

On Line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the OpenCL primitive \verb|get_global_id|.  The type for \verb|gid| is never given explicitly. This is a simple case of Ace's more general {\em whole-function type inference}. In this case, \verb|gid| will be inferred to have type \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification, which the \verb|ace.OpenCL| module follows). The result can be observed on Lines 11 and 24 in Listing \ref{mapout}.

@@

\subsection{Implicit Compilation and Execution}\label{compenv}\label{backend}\label{implicit}
Professional end-users today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}. For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}. Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.\todo{integrate this}

>>>>

The type \verb|TF| on line \ref{compscript}.5 is extracted from the generic function \verb|negate| constructed in Listing \ref{metaprogramming}. Generic functions, according to Sec. \ref{genfn}, have not yet had a type assigned to them, so it may seem perplexing that we are nevertheless extracting a type from it. Although a conventional arrow type cannot be assigned to \verb|negate|, we can give it a \emph{singleton type}: a type that simply means ``this expression is the \emph{particular} generic function \verb|negate|''. This type could also have been explicitly written as \verb|ace.GenericFnType(listing2.negate)|. During typechecking and translation of \verb|map_neg_f32| and \verb|map_neg_ci32|, the call to \verb|f| on line \ref{map}.6 operates by using the types of the provided arguments to compile the generic function that inhabits the singleton type of \verb|f| (\verb|negate| in both of these cases) to a concrete function. This is why there are two versions of \verb|negate| in the output in Listing \ref{mapout}. In other words, types \emph{propagate} into generic functions -- we didn't need to compile \verb|negate| explicitly. This also explains the error printed on line \ref{mapc}.3-\ref{mapc}.4: when this type was inadvertently assigned to the first argument \verb|input|, the indexing operation on line \ref{map}.6 resulted in an error. A generic function can only be \emph{statically} indexed by a list of types to turn it into a concrete function, not \emph{dynamically} indexed with a value of type \verb|clx.size_t| (the return type of the OpenCL primitive function \verb|get_global_id|).

In effect, this scheme enables higher-order functions even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers). Interestingly, because they have a singleton type, they are higher-order but not first-class functions. That is, the type system would prevent you from creating a heterogeneous list of generic functions. Concrete functions, on the other hand, can be given both a singleton type and a true function type. For example, \verb|listing2.scale[[clx.int, clx.int]]| could be given type \verb|ace.Arrow(clx.int, clx.int)|. The base determines how to convert the Ace arrow type to an arrow type in the target language (e.g. a function pointer for C99, or an integer that indexes into a jump table containing all functions in the program of that type).

\todo{move this to related work??}Type assignment to generic functions is similar in some ways to template specialization in C++. In effect, both the template header and type parameters at call sites are being generated automatically by Ace. This significantly simplifies a sophisticated feature of C++ and introduces it to targets like OpenCL and C, which do not support templates. Other uses for C++ templates (and the preprocessor) are subsumed by the metaprogramming features discussed above\todo{cite template metaprogramming paper}.

\subsection{Whole-Function Type Inference}
\begin{codelisting}
\lstinputlisting{listing6.py}
\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single identifier, \texttt{y}.}
\label{inference}
\end{codelisting}
On Line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the OpenCL primitive \verb|get_global_id|.  The type for \verb|gid| is never given explicitly. This is a simple case of Ace's more general {\em whole-function type inference}. In this case, \verb|gid| will be inferred to have type \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification, which the \verb|ace.OpenCL| module follows). The result can be observed on Lines 11 and 24 in Listing \ref{mapout}.

@@

\subsection{Implicit Compilation and Execution}\label{compenv}\label{backend}\label{implicit}
\begin{codelisting}
\lstinputlisting{listing8.py}
\caption{[\texttt{listing\ref{py}.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
\label{py}
\end{codelisting}
Professional end-users today generally use dynamically-typed high-level languages like MATLAB, Python, R or Perl for tasks that are not performance-sensitive, such as small-scale data analysis and plotting \cite{nguyen2010survey}. For portions of their analyses where the performance overhead of dynamic type checking and automatic memory management is too high, they will typically call into code written in a statically-typed, low-level language, most commonly C or Fortran, that uses low-level parallel abstractions like pthreads and MPI \cite{4222616,basili2008understanding}. Unfortunately, these low-level languages and abstractions are notoriously difficult to use and automatic verification is intractable in general.\todo{integrate this}
